Load in required packages:

```{r}
library(expm)
library(dplyr)
library(tidyr)
library(numDeriv)

```

Function to simulate panel data:

```{r}

# simulate MSM according to specified transition intensity matrix Q
msm_sim <- function(q.matrix, max.time = 50, num.subjects = 1) {
  
  # number of possible states
  num.states <- nrow(q.matrix)
  chain.states  <- matrix(0, ncol=num.subjects, nrow=max.time)
  # the first state is assumed to state 1 for subjects
  chain.states[1, ] <- 1
  
  for (subject in 1:num.subjects){
    for(time in 2:max.time) {
      
      # obtain the probability transition matrix at time t
      P  <- expm(time*q.matrix)
      # select the row from P based on the current state
      p  <- P[chain.states[time-1, subject], ]
      
      # draw from multinomial to determine next state
      chain.states[time, subject] <-  which(rmultinom(1, 1, p) == 1)
      
    }
    
  }
  
  # return the simulated data in long form
  data <- chain.states %>% 
    as_tibble() %>% 
    mutate(time = 1:nrow(.)) %>% 
    gather("subject_id", "state", paste0("V", 1:num.subjects)) %>% 
    mutate(subject_id = as.numeric(gsub("V","", subject_id))) %>% 
    select(subject_id, state, time)
  
  
  return(data)
}



```

Function to ensure data is always in a standard format:

```{r}


format_msm_data <- function(msm.data){
  
  # add the lagged/previous state as a column
  # and ensure observations are correctly ordered 
  # by time within each subject
  lagged.msm.data <- msm.data %>% 
    group_by(subject_id) %>% 
    arrange(time) %>% 
    mutate(prev_state = lag(state)) %>% 
    ungroup() %>% 
    # ensures we only have the required columns
    select(
      subject_id, state , time, prev_state
    )
  
  return(lagged.msm.data)
  
  
}

```


Function to compute log likelihood for multi-state model (assuming panel data):

```{r}




log_likelihood <- function(params, msm.data){
  
  n.states <- length(unique(msm.data$state))
  n.subjects <- length(unique(msm.data$subject_id))
  
  # start building up the q matrix
  for(i in seq(1, n.states^2, (n.states+1))){
    
    params <- append(params, NA, after = (i-1))
    
  }
  
  
  q <- matrix(params, n.states, n.states)
  # transpose to get everything in the right order
  q <- t(q)
  
  
  for(i in 1:n.states){
    
    q[i, i] <- -sum(q[i, ], na.rm = TRUE)
    
  }
  
  # compute the probabilties
  p <- expm(q)
  
  # log likelihood calculation
  log_lik <- 0
  
  

  # compute the log likelihood
  # loop over each subject
  for (i in 1:n.subjects){
    
    # subdata frame for subject i
    subject.msm.data <- msm.data %>% 
      filter(subject_id == i)
    
    # number of transitions made 
    n.transitions <- nrow(subject.msm.data)
    
    # loop over each the subject's transitions
    for(j in 2:n.transitions){
      
      index <- unlist(subject.msm.data[, c("prev_state", "state")][j, ], use.names = FALSE)
      
  
      log_lik <- log_lik + log(p[index[1], index[2]])
      
    }
    
  }
    
  return(log_lik)
  
}


```

Define the prior, posterior and proposal functions:

```{r}

# log prior distribution 
log_prior <- function(params){
  
  # number of parameters to estimate
  n.params <- length(params)
  
  # hold the sum of each the priors
  params.prior <- 0
  
  for (i in 1:n.params){
    # assume the priors are weakly informative
    params.prior <- params.prior + dnorm(log(params[i]), sd = 1, log = T)
    
  }
  
  return(params.prior)
}

# log posterior distribution 
log_posterior <- function(params, data){
  
  # evaluate at the exponential since parameters (i.e. param)
  # are sampled on the log scale
  post.dens <- log_likelihood(exp(params), data) + log_prior(exp(params))
  
  return(post.dens)
}



# proposal function 
proposal_values <- function(params, tune_param){
  
  # number of parameters to estimate
  n.params <- length(params)
  
  return(rnorm(n.params, params, tune_param))
}


# Metropolis algorithm 
run_metropolis <- function(start.values, num.samples, tuning.parameter, data){
  
  # make sure data is in a standard format
  mcmc.data <- format_msm_data(msm.data = data)
  # number of parameters to estimate
  n.params <- length(start.values)
  # prepare array to hold samples
  chain = array(dim = c(num.samples+1,n.params))
  chain[1,] = start.values
  
  # set up a progress bar
  pb <- txtProgressBar(min = 1, max = num.samples, style = 3)
  
  for (i in 1:num.samples){
    
  
    
    proposal <- proposal_values(chain[i,], tuning.parameter)
    num <- (log_posterior(proposal, mcmc.data)) 
    denom <- (log_posterior(chain[i,], mcmc.data))
    
    acceptance.ratio <- exp(num - denom)
    
    if (runif(1) < acceptance.ratio){
      chain[i+1,] <- proposal
    }else{
      chain[i+1,] <- chain[i,]
    }
    
    setTxtProgressBar(pb, i)
    
  }
  close(pb)
  return(chain)
}

```

Function to generate plots after Metropolis - Hastings MCMC:

```{r}

mh_mcmc_plots <- function(chain, burn.in){
  n.columns <- ncol(chain)
  
  par(mfrow = c(n.columns,n.columns))
  
  # histograms
  for ( i in 1:n.columns){
    
    hist(chain[-(1:burn.in),i],nclass=30, 
         main=" ",
         xlab = paste0("Log posterior mean: ",
                       round(mean(chain[-(1:burn.in),i]), 2), 
                       "\nParameter estimate: ",
                       
                       round( exp(mean(chain[-(1:burn.in),i])), 2)
                       )
         
         )
    
    abline(v = mean(chain[-(1:burn.in),i]), col = "red")
    
  }
  
  # time series plots 
  for ( i in 1:n.columns){
    
    plot(c(burn.in:iters), chain[-(1:burn.in),i], type = "l",main = " ", ylab = "Simulated value", xlab = "Iteration")
    
  }
  
  par(mfrow = c(1, 1))
  
}

```


Function to generate mesh plot of likelihood surface:

```{r}


plot_likelihood_surface <- function(x.min, x.max, y.min, y.max, data, tick.size, theta){
  
  
  
  msm.data <- format_msm_data(data)
  
  x <- seq(x.min, x.max, tick.size)
  y <- seq(y.min, y.max, tick.size)
  z <- matrix(nrow = length(x), ncol = length(y))
  
  for(i in 1:length(x)){
    for(j in 1:length(y)){
      z[i, j] <- -log_likelihood(params = c(x[i], y[j]), msm.data =  msm.data)
    }
  }
  
  # reset figure rows/columns
  par(mfrow = c(1, 1))
  
 # generate a mesh plot of the likelihood surface
 persp(x, y, z, theta = 110, xlab = "x",
        ylab = "y",
        zlab = "Negative Log-likelihood",
        ticktype = "detailed", 
        main = " ")
  
  
  # plot the likelihood surface heat map
  filled.contour(x = x,y = y,
                 z = z, 
                 color.palette = terrain.colors,
                 plot.title = title(main = " ",
                                    xlab = "x",
                                    ylab = "y"),
                 key.title = {par(cex.main=1);title(main = "Negative\nLog-\nlikelihood")})
  
  message("\nNOTE: Use the plot navigation arrows to view the plots...\n")
  
}




```





--------------------
Examples:
--------------------

Generate some data:
```{r}

# define the q matrix
q <- rbind(
  c(-1.2, 1.2), 
  c(0.8, -0.8)
)

msm.data <- msm_sim(q.matrix = q, num.subjects = 2, max.time = 50)

# check the data
msm.data 

```

Estimate the MSM through MCMM methods:

```{r}
 
# initialisation
tune.param <- 0.6
start.values <- c(1,1)
iters <- 5000
# discard half of the samples for burn in
chain.burn.in <- floor(iters/2)

# run metropolis
log.chain <- run_metropolis(start.values, iters, tune.param, data = msm.data)
acceptance.rate <- 1-mean(duplicated(log_chain[-(1:chain.burn.in),]))

 
```

Summarise the results:

```{r}
 

acceptance.rate
mh_mcmc_plots(log.chain, chain.burn.in)


```

Map out the negative log likelihood surface:

```{r}

plot_likelihood_surface(0.01, 15, 0.01, 15, msm.data, 0.5, 110)

```






